---
title: "Proteomcis in R from a user and developer perspective"
author: "Christian Panse <cp@fgz.ethz.ch> & Tobias Kockmann <tobias.kockmann@fgcz.ethz.ch>"
date: "11/11/2021"
output:
  ioslides_presentation:
    widescreen: yes
    smaller: no
  beamer_presentation: default
---

<style>
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval=TRUE)
source("help-functions.R")
```


## Overview

### Key aspects

* How do we apply R for computational MS (compMS)?
* What have we learned during our own R software development projects?
* Which role will R play in the future compMS landscape?

### Perspectives

* User
* Developer

Disclaimer: Both views are biased by our professional environment

## Our professional environment

The Functional Genomics Center Zurich (FGCZ) is a joint state-of-the-art research and training (core) facility of ETH Zurich and University of Zurich.

[https://fgcz.ch/](https://fgcz.ch/)

We offer/support analytical services, training, collaborative R&D in:

* Genomics
* Metabolomics
* Proteomics

Proteomics currently operates ~12 MS (mostly LC-ESI-MS). Many Orbitraps, fewer TOFs. 

## How we use R

### Tasks

* Generation of sample queues
* System suitability/quality control
* LC-MS method optimization (rawDiag)
* Statisical analysis (MSstats, proLFQa, TPP, ...)
* Reporting (Rmarkdown)

### Modes

* Interactive web applications (Shiny) connected to our core LIMS system
* Scripts are running on HPC cluster as parts of analysis pipelines

## Why we like R... in general 

* R is free and open-source
* R provides an active and engaged community (ask questions, find support)
* R code development is fast
* R code runs on all major hardware platforms (incl. HPC clusters and MC computing)
* R has become the go-to language for statistical modeling/Data Science
* R provides powerful graphical frameworks for data visualization
* Personal history ;-)



## Example R code snippet {.columns-2 .smaller}

for generating an LC-MS instrument queue 

```{r plate}
set.seed(1)

QConf <- 
  iris[c(1:15, 51:65, 101:115), ] |>
  protViz::assignPlatePosition() |>
  protViz::blockRandom(x = "Species") |> 
  protViz::insertSamples(howoften=4,
        stdPosX='6', stdPosY='F', plate=1,
        stdName = "clean",
        volume=2,
        method="C:\\Xcalibur\\general_clean") 

```

<p class="forceBreak"></p>

```{r plotPlate, fig.width=4.5, fig.height=5, echo=FALSE}
op <- par(mar=c(4,4,1,1))
  .plotPlate(QConf, 'Species')
```


## Lessons learned




## What can R achieve?

* AI?




# BACKUP

## R interaction with a typical proteomics pipeline



## What are we providing?
Service, training, tailored R&D in the area of:

- Genomics
- Metabolomics
- Proteomics

## Orbitrap run statistics
As core facility we generate ~20K Orbitrap runs a year:

```{r numberRuns, fig.retina=3, echo=FALSE}
S <- data.frame(year = c(2012:2020),
           all = c(12295, 15982, 15190, 16726, 19251, 24588, 28814, 30531, 27176),
           clean = c(881, 1788, 1700,  2228, 2121, 2965, 4097, 6677, 6529),
           qc = c(1952,  3326, 2542,2821, 4027, 4435, 5640, 5247, 3761))
plot(all ~ year, data=S, pch=16, type='b', lwd=2, ylim=c(0,max(S$all)), xlim=c(2012,2020), ylab='number of runs'); lines(S$qc ~ S$year,type='b', col='red'); lines(S$clean ~ S$year,type='b', col='green'); legend("topleft", c('all', 'qc', 'clean'), pch=16, col=c('black', 'red', 'green')); box(); lines(S$all-(S$qc +S$clean) ~ S$year,type='b', col='blue');
```

while operating up to 15 Orbitrap based LC-MS systems. (strong ThermoFisher MS fleet)

<div class='notes'>
|year  |  2012|  2013|  2014|  2015|  2016|  2017|  2018|  2019|  2020| 2021|
|:-----|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|----:|
|all   | 12295| 15982| 15190| 16726| 19251| 24588| 28814| 30531| 27176| 6261|
|clean |   881|  1788|  1700|  2228|  2121|  2965|  4097|  6677|  6529| 1449|
|qc    |  1952|  3326|  2542|  2821|  4027|  4435|  5640|  5247|  3761|  820|

</div>


## How do we use R in our core facility?

- Quality control 
- LC-MS method optimization
- Statistics and visualization
- Automatic report generation

## Example on how we use R

- autoQC
- rawDiag
- reporting
- proLFQa

### DISCLAIMER
In **proteomics**
we do **NOT** use R for identification and quantification
(no LC-MS feature finding) tasks.

<div class="note">
here we use mascot, comet, MaxQuant, MS-FRAGGER, Spectronaut ...
</div>




## Idea software

- runs on multi-process/user systems on all major platforms
- software ship with support 
- license is independent of the number of users



# Related work

## Todays common uses software pipelines in proteomics

MaxQuant
FragPipe
Biognosys
...

## R related proteomics packages

CRAN:
 https://CRAN.R-project.org/package=opentimsr 
 https://CRAN.R-project.org/package=protViz
 
Bioconductor:
 https://bioconductor.org/packages/release/BiocViews.html#___Proteomics
 
## R package frameworks

Spectra


