---
title: "Proteomcis in R"
author: "Christian Panse <cp@fgz.ethz.ch> & Tobias Kockmann <tobias.kockmann@fgcz.ethz.ch>"
date: "11/11/2021"
output:
  ioslides_presentation:
    widescreen: yes
    smaller: no
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval=TRUE)
```

## What are we providing?
Service, training, tailored R&D in the area of:

- Genomics
- Metabolomics
- Proteomics

## Orbitrap run statistics
As core facility we generate ~20K Orbitrap runs a year:

```{r numberRuns, fig.retina=3, echo=FALSE}
S <- data.frame(year = c(2012:2020),
           all = c(12295, 15982, 15190, 16726, 19251, 24588, 28814, 30531, 27176),
           clean = c(881, 1788, 1700,  2228, 2121, 2965, 4097, 6677, 6529),
           qc = c(1952,  3326, 2542,2821, 4027, 4435, 5640, 5247, 3761))
plot(all ~ year, data=S, pch=16, type='b', lwd=2, ylim=c(0,max(S$all)), xlim=c(2012,2020), ylab='number of runs'); lines(S$qc ~ S$year,type='b', col='red'); lines(S$clean ~ S$year,type='b', col='green'); legend("topleft", c('all', 'qc', 'clean'), pch=16, col=c('black', 'red', 'green')); box(); lines(S$all-(S$qc +S$clean) ~ S$year,type='b', col='blue');
```

while operating up to 15 Orbitrap based LC-MS systems. (strong ThermoFisher MS fleet)

<div class='notes'>
|year  |  2012|  2013|  2014|  2015|  2016|  2017|  2018|  2019|  2020| 2021|
|:-----|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|-----:|----:|
|all   | 12295| 15982| 15190| 16726| 19251| 24588| 28814| 30531| 27176| 6261|
|clean |   881|  1788|  1700|  2228|  2121|  2965|  4097|  6677|  6529| 1449|
|qc    |  1952|  3326|  2542|  2821|  4027|  4435|  5640|  5247|  3761|  820|

</div>


## How do we use R in our core facility?

- Quality control 
- LC-MS method optimization
- Statistics and visualization
- Automatic report generation

## Example on how we use R

- autoQC
- rawDiag
- reporting
- proLFQa

### DISCLAIMER
In **proteomics**
we do **NOT** use R for identification and quantification
(no LC-MS feature finding) tasks.

<div class="note">
here we use mascot, comet, MaxQuant, MS-FRAGGER, Spectronaut ...
</div>




## Idea software

- runs on multi-process/user systems on all major platforms
- software ship with support 
- license is independent of the number of users


## Lessons learned

...
...


# Related work

## Todays common uses software pipelines in proteomics

MaxQuant
FragPipe
Biognosys
...

## R related proteomics packages

CRAN:
 https://CRAN.R-project.org/package=opentimsr 
 https://CRAN.R-project.org/package=protViz
 
Bioconductor:
 https://bioconductor.org/packages/release/BiocViews.html#___Proteomics
 
## R package frameworks

Spectra
